---
title: "Approximating Palettes from Other Packages"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: color.bib
vignette: >
  %\VignetteIndexEntry{Approximating Palettes from Other Packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{colorspace,RColorBrewer,rcartocolor,viridis}
  %\VignetteKeywords{HCL, palettes, spectrum, ColorBrewer.org, CARTO, viridis}
  %\VignettePackage{colorspace}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("colorspace")
library("RColorBrewer")
library("rcartocolor")
library("viridis")
library("scico")
```

## Overview

The flexible specification of HCL-based color palettes in _colorspace_ allows
to closely approximate color palettes from various other packages:

- _ColorBrewer.org_ [@color:Harrower+Brewer:2003] as provided by R package _RColorBrewer_ [@color:RColorBrewer].
- _CARTO_ colors [@color:CARTO] as provided by R package _rcartocolor_ [@color:rcartocolor].
- The viridis palettes of @color:Smith+VanDerWalt:2015 developed for _matplotlib_,
  as provided by R package _viridis_ [@color:viridis].
- The scientific color maps of @color:Crameri:2018 as provided by R package _scico_ [@color:scico].

See the discussion of [HCL-based palettes](hcl_palettes.html) for more details. In the
following `specplot()` is used to compare the HCL spectrum of the original palettes (top swatch, solid lines)
and their HCL-based approximations (bottom swatch, dashed lines).


## Approximations of ColorBrewer.org palettes

```{r brewer, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, dev = "png", dpi = 48}
demo("brewer", package = "colorspace")
```


## Approximations of CARTO palettes

```{r carto, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, dev = "png", dpi = 48}
demo("carto", package = "colorspace")
```


## Approximations of viridis palettes

```{r viridis, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, dev = "png", dpi = 48}
demo("viridis", package = "colorspace")
```


## Approximations of Crameri's scientific color (scico) palettes

```{r scico, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, dev = "png", dpi = 48}
demo("scico", package = "colorspace")
```


## References
