---
title: "Approximating Palettes from Other Packages"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: color.bib
vignette: >
  %\VignetteIndexEntry{Approximating Palettes from Other Packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{colorspace,RColorBrewer,rcartocolor,viridis}
  %\VignetteKeywords{HCL, palettes, spectrum, ColorBrewer.org, CARTO, viridis}
  %\VignettePackage{colorspace}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("colorspace")
library("RColorBrewer")
library("rcartocolor")
library("viridis")
library("scico")
```

## Overview

The flexible specification of HCL-based color palettes in _colorspace_ allows
to closely approximate color palettes from various other packages:

* _ColorBrewer.org_ [@color:Harrower+Brewer:2003] as provided by R package _RColorBrewer_ [@color:RColorBrewer].
* _CARTO_ colors [@color:CARTO] as provided by R package _rcartocolor_ [@color:rcartocolor].
* The viridis palettes of @color:Smith+VanDerWalt:2015 developed for _matplotlib_,
  as provided by R package _viridis_ [@color:viridis].
* The scientific color maps of @color:Crameri:2018 as provided by R package _scico_ [@color:scico].

See the discussion of [HCL-based palettes](hcl_palettes.html) for more details. In the
following sections `specplot()` is used to compare the [HCL spectrum](palette_visualization.html#hcl-and-rgb-spectrum)
of the original palettes (top swatches, solid lines) and their HCL-based approximations (bottom swatches, dashed lines).

Before, a selection of such approximations using `specplot()` is highlighted and discussed in some more detail.
Specifically, the graphic below shows two blue/green/yellow palettes (`RColorBrewer::brewer.pal(7, "YlGnBu")`
and `viridis::viridis(7)`) and two purple/red/yellow (`rcartocolor::carto_pal(7, "ag_Sunset")` and `viridis::plasma(7)`).
Each panel compares the hue, chroma, and luminance trajectories
of the original palettes (top swatches, solid lines) and their HCL-based approximations (bottom swatches, dashed lines).
The palettes are not identical but very close for most colors. Note that also the chroma trajectories from the HCL
palettes (green dashed lines) have some kinks which are due to fixing HCL coordinates at the boundaries of
admissible RGB colors.

```{r brewer-carto-viridis, echo = FALSE, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, dev = "png", dpi = 48}
specplot(
  rev(RColorBrewer::brewer.pal(7, "YlGnBu")),
  sequential_hcl(7, "YlGnBu"),
  main = "ColorBrewer.org: YlGnBu"
)
specplot(
  viridis::viridis(7),
  sequential_hcl(7, "Viridis"),
  main = "viridis: Viridis"
)
specplot(
  rcartocolor::carto_pal(7, "ag_Sunset"),
  sequential_hcl(7, "ag_Sunset"),
  main = "CARTO: ag_Sunset"
)
specplot(
  viridis::plasma(7),
  sequential_hcl(7, "Plasma"),
  main = "viridis: Plasma"
)
```

These graphics illustrate what sets the viridis palettes apart from
other sequential palettes. While the hue and luminance trajectories of `"Viridis"` and `"YlGnBu"` are very similar,
the chroma trajectories differ: While lighter colors (with high luminance) have low chroma for `"YlGnBu"`, they have
increasing chroma for `"Viridis"`. Similarly, `"ag_Sunset"` and `"Plasma"` have similar hue and luminance trajectories
but different chroma trajectories. The result is that the viridis palettes have rather high chroma throughout which
does not work so well for sequential palettes on a white/light background as all shaded areas convey high "intensity".
However, they work better on a dark/black background. Also, they might be a reasonable
alternative for qualitative palettes when grayscale printing should also work.

Another somewhat nonstandard palette from the viridis family is the
cividis palette based on blue and yellow hues and hence safe for red-green deficient viewers.
The figure below shows the corresponding `specplot()` along with an HCL-based approximation.
What is unusual about this palette: The hue and chroma trajectories would suggest a diverging palette,
as there are two "arms" wth different hues and a zero-chroma point in the center. However,
the luminance trajectory clearly indicates a sequential palette as colors go monotonically from
dark to light. Due to this unusual mixture the palette cannot be composed using the trajectories
discussed in the [construction details](hcl_palettes.html#construction-details).

```{r cividis, echo = FALSE, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, dev = "png", dpi = 48}
specplot(viridis::cividis(9), divergingx_hcl(9, "Cividis"), main = "viridis: Cividis")
```

However, the tools in _colorspace_ can still be employed to easily reconstruct the palette.
One strategy would be to set up the trajectories manually, using a linear luminance, piecewise
linear chroma, and piecewise constant hue:

```{r cividis_hcl}
cividis_hcl <- function(n) {
  i <- seq(1, 0, length.out = n)
  hex(polarLUV(
    L = 92 - (92 - 13) * i,
    C = approx(c(1, 0.9, 0.5, 0), c(30, 50, 0, 95), xout = i)$y,
    H = c(255, 75)[1 + (i < 0.5)]
  ), fix = TRUE)
}
```

Instead of constructing the hex code from the HCL coordinates via _colorspace_'s `hex(polarLUV(L, C, H))`,
the base R function `hcl(H, C, L)` from _grDevices_ could also be used.

In addition to manually setting up a dedicated function `cividis_hcl()`, it is possible to approximate the
palette using `divergingx_hcl()`, e.g.,

```{r diverging_cividis, eval = FALSE}
divergingx_hcl(n,
  h1 = 255, h2 = NA, h3 = 75,
  c1 = 30, cmax1 = 47, c2 = 0, c3 = 95,
  l1 = 13, l2 = 52, l3 = 92,
  p1 = 1.1, p3 = 1.0
)
```

This uses a slight power transformation with `p1 = 1` in the blue arm of the palette but otherwise essentially
corresponds to what `cividis_hcl()` does. For convenience `divergingx_hcl(n, palette = "Cividis")` is preregistered
using the above parameters.


## Approximations of ColorBrewer.org palettes

```{r brewer, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, dev = "png", dpi = 48}
demo("brewer", package = "colorspace")
```


## Approximations of CARTO palettes

```{r carto, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, dev = "png", dpi = 48}
demo("carto", package = "colorspace")
```


## Approximations of viridis palettes

```{r viridis, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, dev = "png", dpi = 48}
demo("viridis", package = "colorspace")
```


## Approximations of Crameri's scientific color (scico) palettes

```{r scico, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, dev = "png", dpi = 48}
demo("scico", package = "colorspace")
```


## References
