---
title: "Color Spaces: S4 Classes and Utilities"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: color.bib
vignette: >
  %\VignetteIndexEntry{Color Spaces: S4 Classes and Utilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{colorspace}
  %\VignetteKeywords{RGB, sRGB, XYZ, LUV, LAB, HLS, HSV, HCL}
  %\VignettePackage{colorspace}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("colorspace")
```

## Overview

At the core of the _colorspace_ package there are various utilities for computing with
color spaces (as the name conveys). Thus, the package helps to map various three-dimensional
representations of color to each other, see @color:Ihaka:2003. A particularly important
mapping was the one from the perceptually-based and device-independent color model
HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
in R). For completeness further standard color models were included as well in the package.
The connections are illustrated in the following graph:


```{r color-spaces, fig.width = 9.5, fig.height = 4.5, fig.align = "center", echo = FALSE, dev = "png"}
par(mar = rep(0, 4))

d <- data.frame(
  x = c(1, 1, 2, 2, 3, 4, 5, 5, 5, 6),
  y = c(1, 3, 1, 3, 2, 2, 1, 3, 2, 2),
  lab = c("polarLAB", "polarLUV\n(= HCL)", "LAB", "LUV", "XYZ", "RGB", "HLS", "HSV", "sRGB", "hex"),
  stringsAsFactors = FALSE
)

plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
  xlim = c(0.7, 6.3), ylim = c(0.4, 3.6))
lines(y ~ x, data = d[-c(1, 3, 7, 8),])
lines(y ~ x, data = d[c(1, 3, 5),])
lines(y ~ x, data = d[c(6, 7, 9),])
lines(y ~ x, data = d[c(6, 8, 9),])
points(y ~ x, data = d[c(1, 2, 7, 8), ], pch = 21, bg = "lightgray", cex = 12)
points(y ~ x, data = d[c(3, 4, 5, 6, 9), ], pch = 22, bg = "lightgray", cex = 12)
points(y ~ x, data = d[10, ], pch = 23, bg = "white", cex = 12)

text(d$x, d$y, d$lab)
text(3.5, 2, "white point\n= D65", cex = 0.7, pos = 3)
text(4.56, 2, "gamma\n(= 2.4)", cex = 0.7, pos = 3)
```

Color models that are (or try to be) perceptually-based are displayed with circles,
other color models with rectangles. The corresponding classes and eponymous class
constructors in _colorspace_ are:

* `RGB()` for the classic Red-Green-Blue color model that mixes three primary colors
  with different intensities to obtain a spectrum of colors. The advantage of this
  color model is (or was) that it corresponded to how computer and TV screens generated
  colors, hence it was widely adopted and still is the basis for color specifications
  in many systems. For example, the hex color codes are employed in HTML but also in R.
  However, the RGB model also has some important drawbacks: It does not take into
  account the output device properties, it is not perceptually uniform (a unit step
  within RGB does not produce a constant perceptual change in color), and it is
  unintuitive for humans to specify colors (say brown or pink) in this space.
  [@color:Wiki+RGB]
* `sRGB()` addresses the issue of device dependency by adopting a so-called gamma
  correction. Therefore, the gamma-corrected standard RGB (sRGB), as opposed to the
  linearized RGB above, is a good model for specifying colors in software and for
  hardware. But it is still unintuitive for humans to work directly with this.
  Therefore, sRGB is a good place to end up in a color space manipulation but
  it is not a good place to start. [@color:Wiki+sRGB:2018]
* `HSV()` is a simple transformation of the (s)RGB space that tries to capture the
  perceptual axes: _hue_ (dominant wavelength, the type of color), _saturation_
  (colorfulness), and _value_ (brightness, i.e., light vs. dark). Unfortunately,
  the three axes in the HSV model are confounded so that, e.g., brightness changes
  dramaticaly with hue.
  [@color:Wiki+HSV]
* `HLS()` (Hue-Lightness-Saturation) is another transformation of (s)RGB that tries
  to capture the perceptual axes. It does a somewhat better job but the dimensions
  are still strongly confounded.
  [@color:Wiki+HSV]
* `XYZ()` was established by the CIE (Commission Internationale de l'Eclairage)
  based on experiments with human subjects. It provides a unique triplet of
  XYZ values, coding the standard observer's perception of the color. It is
  device-independent but it is not perceptually uniform and the XYZ coordinates
  have no intuitive meaning.
  [@color:Wiki+CIEXYZ]
* `LUV()` and `LAB()` were therefore proposed by the CIE as perceptually uniform
  color spaces where the former is typically preferred for emissive technologies
  (such as screens and monitors) whereas the latter is usually preferred when
  working with dyes and pigments. However, the three axes of these two spaces
  still do not correspond to human perceptual axes.
  [@color:Wiki+CIELUV; @color:Wiki+CIELAB]
* `polarLUV()` and `polarLAB()` therefore take polar coordinates in the UV plane
  and AB plane, respectively. Specifically, the polar coordinates of the LUV
  model are known as the HCL (Hue-Chroma-Luminance) model. These capture the
  human perceptual axes very well without confounding effects as in the HSV or
  HLS approaches. (More details follow below.)

All the S4 classes listed above inherit from a virtual class `color` which are
internally always represented by matrices with three columns (corresponding to
the different three dimensions).


## Utilities

For working with these color spaces various utilities are available:

* `as()` method: Conversions of a `color` object to the various color spaces,
  e.g., `as(x, "sRGB").
* `coords()`: Extract the three-dimensional coordinates pertaining to the current
  color class.
* `hex()`: Convert a color object to sRGB and code in a hex string that can be
  used within R plotting functions.
* `hex2RGB()`: Convert a given hex color string to an sRGB color object which
  can also be coerced to other color spaces.
* `readRGB()` and `readhex()` can read text files into color objects, either
  from RGB coordinates or hex color strings.
* `writehex()`: Writes hex color strings to a text file.
* `whitepoint()`: Query and change the white point employed in conversions
  from CIE XYZ to RGB. Defaults to D65.


## Illustration of basic colorspace functionality

As an example a vector of colors `x` can be specified in the HCL (or polar LUV) model:

```{r polarLUV}
(x <- polarLUV(L = 70, C = 50, H = c(0, 120, 240)))
```

These are pastel red (hue = 0), green (hue = 120), and blue (hue = 240) with moderate chroma and luminance.
For display in other systems an sRGB representation might be needed:

```{r sRGB}
(y <- as(x, "sRGB"))
```

With `coords(x)` or `coords(y)` the displayed coordinates can also be extracted
as numeric matrices. And from sRGB we can also coerce to HSV for example:

```{r HSV}
as(y, "HSV")
```

For display in many systems (including R itself) hex color codes based on the
sRGB coordinates can be created:

```{r hex}
hex(x)
```

## References
