---
title: "Color Spaces: S4 Classes and Utilities"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: color.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Color Spaces: S4 Classes and Utilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{colorspace}
  %\VignetteKeywords{RGB, sRGB, XYZ, LUV, LAB, HLS, HSV, HCL}
  %\VignettePackage{colorspace}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("colorspace")
```

## Overview

At the core of the _colorspace_ package are various utilities for computing with
color spaces [@color:Wiki+Colorspace], as the name conveys. Thus, the package helps to map various three-dimensional
representations of color to each other [@color:Ihaka:2003]. A particularly important
mapping is the one from the perceptually-based and device-independent color model
HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
specifications in many systems based on the corresponding hex codes [@color:Wiki+Webcolors],
e.g., in HTML but also in R. For completeness further standard color models are included as well in the package.
Their connections are illustrated in the following graph:


```{r color-spaces, fig.width = 9.5, fig.height = 4.5, fig.align = "center", echo = FALSE, dev = "png"}
par(mar = rep(0, 4))

d <- data.frame(
  x = c(1, 1, 2, 2, 3, 4, 5, 5, 5, 6),
  y = c(1, 3, 1, 3, 2, 2, 1, 3, 2, 2),
  lab = c("polarLAB", "polarLUV\n(HCL)", "LAB", "LUV", "XYZ", "RGB", "HLS", "HSV", "sRGB", "hex"),
  stringsAsFactors = FALSE
)

plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
  xlim = c(0.7, 6.3), ylim = c(0.4, 3.6))
lines(y ~ x, data = d[-c(1, 3, 7, 8),])
lines(y ~ x, data = d[c(1, 3, 5),])
lines(y ~ x, data = d[c(6, 7, 9),])
lines(y ~ x, data = d[c(6, 8, 9),])
points(y ~ x, data = d[c(1, 2, 7, 8), ], pch = 21, bg = "lightgray", cex = 12)
points(y ~ x, data = d[c(3, 4, 5, 6, 9), ], pch = 22, bg = "lightgray", cex = 12)
points(y ~ x, data = d[10, ], pch = 23, bg = "white", cex = 12)

text(d$x, d$y, d$lab)
text(3.5, 2, "white point\n= D65", cex = 0.7, pos = 3)
text(4.56, 2, "gamma\n= 2.4", cex = 0.7, pos = 3)
```

Color models that are (or try to be) perceptually-based are displayed with circles
and models that are not are displayed with rectangles. The corresponding classes and eponymous class
constructors in _colorspace_ are:

* `RGB()` for the classic Red-Green-Blue color model, which mixes three primary colors
  with different intensities to obtain a spectrum of colors. The advantage of this
  color model is (or was) that it corresponded to how computer and TV screens generated
  colors, hence it was widely adopted and still is the basis for color specifications
  in many systems. For example, hex color codes are employed in HTML but also in R.
  However, the RGB model also has some important drawbacks: It does not take into
  account the output device properties, it is not perceptually uniform (a unit step
  within RGB does not produce a constant perceptual change in color), and it is
  unintuitive for humans to specify colors (say brown or pink) in this space.
  [See @color:Wiki+RGB.]
* `sRGB()` addresses the issue of device dependency by adopting a so-called gamma
  correction. Therefore, the gamma-corrected standard RGB (sRGB), as opposed to the
  linearized RGB above, is a good model for specifying colors in software and for
  hardware. But it is still unintuitive for humans to work directly with this color space.
  Therefore, sRGB is a good place to end up in a color space manipulation but
  it is not a good place to start. [See @color:Wiki+sRGB.]
* `HSV()` is a simple transformation of either the sRGB or the RGB space that tries to capture the
  perceptual axes: _hue_ (dominant wavelength, the type of color), _saturation_
  (colorfulness), and _value_ (brightness, i.e., light vs. dark). Unfortunately,
  the three axes in the HSV model are confounded so that, e.g., brightness changes
  dramatically with hue.
  [See @color:Wiki+HSV.]
* `HLS()` (Hue-Lightness-Saturation) is another transformation of either sRGB or RGB that tries
  to capture the perceptual axes. It does a somewhat better job but the dimensions
  are still strongly confounded.
  [See @color:Wiki+HSV.]
* `XYZ()` was established by the CIE (Commission Internationale de l'Eclairage)
  based on experiments with human subjects. It provides a unique triplet of
  XYZ values, coding the standard observer's perception of the color. It is
  device-independent but it is not perceptually uniform and the XYZ coordinates
  have no intuitive meaning.
  [See @color:Wiki+CIEXYZ.]
* `LUV()` and `LAB()` were therefore proposed by the CIE as perceptually uniform
  color spaces where the former is typically preferred for emissive technologies
  (such as screens and monitors) whereas the latter is usually preferred when
  working with dyes and pigments. The L coordinate in both spaces has the same
  meaning and captures luminace (light-dark contrasts). Both the U and V coordinates
  as well as the A and B coordinates measure positions on red/green and yellow/blue
  axes, respectively, albeit in somewhat different ways. While this corresponds
  to how human color vision likely evolved (see the next section), these two color
  models still not correspond to perceptual axes that humans use to describe colors.
  [See also @color:Wiki+CIELUV; @color:Wiki+CIELAB.]
* `polarLUV()` and `polarLAB()` therefore take polar coordinates in the UV plane
  and AB plane, respectively. Specifically, the polar coordinates of the LUV
  model are known as the HCL (Hue-Chroma-Luminance) model [see @color:Wiki+HCL, 
  which points out that the LAB-based polar coordinates are also sometimes referred to as HCL].
  The HCL model captures the human perceptual axes very well without confounding
  effects as in the HSV or HLS approaches. (More details follow below.)

All S4 classes for color spaces inherit from a virtual class `color` which is
internally always represented by matrices with three columns (corresponding to
the three dimensions).

Note that since the inception of the color space conversion tools in \pkg{colorspace}
[in \proglang{C}, @color:Ihaka:2003] other \proglang{R} tools for this purpose became available,
notably `grDevices::convertColor()` [in high-level \proglang{R}, @color:R] and
`farver::convert_colour()` [in \proglang{C++}, @color:farver]. For many basic color
conversion purposes the _colorspace_ package and these alternatives are equally suitable.
Some details differ, though, and the color conversion tools in _colorspace_ are
well-embedded into the remaining functionality of the package and make it self-contained.


## Human color vision and the HCL color model

It has been hypothesized that human color vision has evolved in three distinct stages:

1. Perception of light/dark contrasts (monochrome only).
2. Yellow/blue contrasts (usually associated with our notion of warm/cold colors).
3. Green/red contrasts (helpful for assessing the ripeness of fruit).

See @color:Kaiser+Boynton:1996, @color:Knoblauch:2002, @color:Ihaka:2003, @color:Lumley:2006, @color:Zeileis+Hornik+Murrell:2009
for more details and references. Thus, colors can be described using a 3-dimensional space:

```{r human-axes, echo = FALSE, results = "hide"}
grDevices::svg(file = "screenshots/human-axes.svg", width = 4.5, height = 4.5)
par(mar = rep(0, 4))
midgray <- gray(134/255)
res <- persp(c(-1, 1), c(-1, 1), matrix(c(-1, 1, 1, -1), 2, 2),
  xlim = c(-0.77, 0.77), ylim = c(-0.77, 0.77), zlim = c(-0.77, 0.77),
  border = "transparent", box = FALSE, axes = FALSE, phi = 20, theta = 30, r = 50)
ax <- trans3d(c(1, -1, 0, 0, 0, 0), c(0, 0, 1, -1, 0, 0), c(0, 0, 0, 0, 1, -1), res)
circ <- (-45:45)/50
circ <- trans3d(rep(circ, 2), c(sqrt(circ[1]^2 - circ^2), -sqrt(circ[1]^2 - circ^2)), 0, res)
arrows(ax$x[c(1, 3, 5)], ax$y[c(1, 3, 5)], ax$x[c(2, 4, 6)], ax$y[c(2, 4, 6)], code = 3, length = 0.1, col = midgray, lwd = 3)
lines(circ$x, circ$y, col = midgray, lty = 2)
text(ax$x, ax$y, c("green", "red", "blue", "yellow", "light", "dark"), col = midgray, pos = c(4, 2, 4, 2, 3, 1))
dev.off()
```

[![](screenshots/human-axes.svg)](screenshots/human-axes.svg)

However, for describing colors in such a space, it is more natural for humans to employ
polar coordinates in the color plane (yellow/blue vs. green/red, visualized by the dashed circle above)
plus a third light/dark axis.
Hence, color models that attempt to capture these perceptual axes are also called perceptually-based color spaces.
As already argued above, the HCL model captures these dimensions very well, calling them: _hue_, _chroma_, and _luminance_.
The corresponding sRGB gamut, i.e., the HCL colors that can also be represented in sRGB, is visualized
in the animation below [by @color:Horvath+Lipka:2017].

<iframe src="https://commons.wikimedia.org/wiki/File:SRGB_gamut_within_CIELCHuv_color_space_mesh.webm?embedplayer=yes" width="512" height="512" frameborder="0" ></iframe>

The shape of the HCL space is a distorted double cone which is seen best by looking at
vertical slices, i.e., chroma-luminance planes for given hues. For example, the left panel below depicts
the chroma-luminance plane for a certain blue (hue = 255). Along with luminance the colors change from
dark to light. With increasing chroma the colors become more colorful, where the highest chroma is
possible for intermediate luminance.

As some colors are relatively dark (e.g., blue and red assume their maximum chroma for relatively low luminances)
while others are relatively light (e.g., yellow and green), horizontal slices of hue-chroma planes for given hue
have somewhat irregular shapes. The right panel below shows such a hue-chroma plane for moderately light colors
(luminance = 70). At that luminance, green and orange can become much more colorful compared to blue or red.

```{r hcl-projections, fig.width = 9.5, fig.height = 4.5, fig.align = "center", echo = FALSE, dev = "png"}
par(mfrow = c(1, 2))
hclplot(sequential_hcl(5), h = 255, c = 125, size = 0, lwd = 0)
hclplot(qualitative_hcl(5), l = 70, c = 125, size = 0, lwd = 0)
```

## Utilities

Several utilities are available for working with the S4 classes implementing the
color spaces listed above.

* `as()` method: Convert a `color` object to the various color spaces,
  e.g., `as(x, "sRGB")`.
* `coords()`: Extract the three-dimensional coordinates pertaining to the current
  color class.
* `hex()`: Convert a color object to sRGB and code in a hex string that can be
  used within R plotting functions.
* `hex2RGB()`: Convert a given hex color string to an sRGB color object which
  can also be coerced to other color spaces.
* `readRGB()` and `readhex()` can read text files into color objects, either
  from RGB coordinates or hex color strings.
* `writehex()`: Write hex color strings to a text file.
* `whitepoint()`: Query and change the so-called white point employed in conversions
  from CIE XYZ to RGB. Defaults to D65 that has been specified by the CIE to
  approximate daylight.


## Illustration of basic colorspace functionality

As an example a vector of colors `x` can be specified in the HCL (or polar LUV) model:

```{r polarLUV}
(x <- polarLUV(L = 70, C = 50, H = c(0, 120, 240)))
```

The resulting three colors are pastel red (hue = 0), green (hue = 120), and blue (hue = 240) with moderate chroma and luminance.
For display in other systems an sRGB representation might be needed:

```{r sRGB}
(y <- as(x, "sRGB"))
```

The displayed coordinates can also be extracted as numeric matrices by
`coords(x)` or `coords(y)`. We can also, for example, coerce from sRGB to HSV:

```{r HSV}
as(y, "HSV")
```

For display in many systems (including R itself) hex color codes based on the
sRGB coordinates can be created:

```{r hex}
hex(x)
```

## References
