---
title: "_colorspace_: A Toolbox for Manipulating and Assessing Colors and Palettes"
author: "Achim Zeileis, Jason C. Fisher, Kurt Hornik, Ross Ihaka, Claire D. McWhite, Paul Murrell, Reto Stauffer, Claus O. Wilke"
date: 2018-09-20
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
bibliography: color.bib
vignette: >
  %\VignetteIndexEntry{colorspace: A Toolbox for Manipulating and Assessing Colors and Palettes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{colorspace}
  %\VignetteKeywords{RGB, sRGB, XYZ, LUV, LAB, HLS, HSV, HCL, qualitative palette, sequential palette, diverging palette, shiny, visualization, color vision deficiency}
  %\VignettePackage{colorspace}
---
```{r echo=FALSE, message=FALSE}
library("colorspace")
```

# Color spaces

At the core of the _colorspace_ package there are various utilities for computing with
color spaces (as the name conveys). Thus, the package helps to map various three-dimensional
representations of color to each other, see @color:Ihaka:2003. A particularly important
mapping was the one from the perceptually-based and device-independent color model
HCL (Hue-Chroma-Luminance) to standard Red-Green-Blue (sRGB) which is the basis for color
specifications in many systems based on the corresponding hex codes (e.g., in HTML but also
in R). For completeness further standard color models were included as well in the package:
`polarLUV()` (= HCL), `LUV()`, `polarLAB()`, `LAB()`, `XYZ()`, `RGB()`, `sRGB()`, `HLS()`,
`HSV()`.

For working with these color spaces various utilities are available, including
conversions to other color space representations or to hex color codes etc.

**More details:** [Color Spaces: S4 Classes and Utilities](color_spaces.html).


# HCL-based palettes

The HCL space (= polar coordinates in CIELUV) is particularly useful for
specifying individual colors and color palettes as its three axes match those
of the human visual system very well. Therefore, the _colorspace_ package provides
three types of palettes based on the HCL model:

* _Qualitative:_ Designed for coding categorical information, i.e.,
  where no particular ordering of categories is available and every color
  should receive the same perceptual weight.
* _Sequential:_ Designed for coding ordered/numeric information, i.e.,
  where colors go from high to low (or vice versa).
* _Diverging:_ Designed for coding numeric information around a central
  neutral value, i.e., where colors diverge from neutral to two extremes.

The corresponding functions are `qualitative_hcl()`, `sequential_hcl()`,
and `diverging_hcl()`. 

**More details:** [`hcl_palettes`: HCL-Based Color Palettes](hcl_palettes.html).


# Palette visualization and assessment


## Color swatches: `swatchplot()`

swatches of several palette vectors

```{r fig.width = 4, fig.height = 2.2, fig.align = "center", dev = "png"}
swatchplot(
  "Hue"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
  "Chroma"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
  "Luminance" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
  off = 0
)
```

swatches of named palette matrices

```{r fig.width = 11.5, fig.height = 3.5, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
bprg <- c("Blues", "Purples", "Reds", "Greens")
swatchplot(
  "Single-hue"            = t(sapply(paste(bprg, 2), sequential_hcl, n = 7)),
  "Single-hue (advanced)" = t(sapply(paste(bprg, 3), sequential_hcl, n = 7)),
  "Multi-hue (advanced)"  = t(sapply(bprg,           sequential_hcl, n = 7)),
  nrow = 5
)
```

## HCL (and RGB) spectrum: `specplot()`

spectrum of HCL-based palettes: qualitative/sequential/diverging

```{r fig.width = 5, fig.height = 5, fig.align = "center", dev = "png"}
specplot(qualitative_hcl(100, "Set 2"))
```

```{r fig.width = 5, fig.height = 5, fig.align = "center", dev = "png"}
specplot(sequential_hcl(100, "Blues 2"))
```

```{r fig.width = 5, fig.height = 5, fig.align = "center", dev = "png"}
specplot(diverging_hcl(100, "Blue-Red"))
```

spectrum of the (in)famous RGB rainbow palette (in both RGB and HCL)

```{r fig.width = 5, fig.height = 7.5, fig.align = "center", dev = "png"}
specplot(rainbow(100), rgb = TRUE)
```


## Trajectories in HCL space: `hclplot()`

For qualitative palettes luminance and chroma are fixed, varying only hue

```{r fig.width = 4, fig.height = 4, fig.align = "center", dev = "png"}
hclplot(qualitative_hcl(9, c = 50, l = 70))
```

Single-hue sequential palette (h = 260) with linear trajectory and
advanced single-hue sequential palette with triangular chroma trajectory
(piecewise linear vs. power-transformed).

```{r fig.width = 11.5, fig.height = 4.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = 260, c = 80, l = c(35, 95), power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95), power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95), power = c(0.8, 1.4)))
```

Multi-hue sequential palette with small hue range and triangular chroma vs.
large hue range and linear chroma trajectory.

Balanced diverging palette constructed from two simple single-hue sequential
palettes (for hues 260/blue and 0/red)

```{r fig.width = 11.5, fig.height = 4.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = c(260, 220), c = c(50, 75, 0), l = c(30, 95), power = 1))
hclplot(sequential_hcl(7, h = c(260, 60), c = 60, l = c(40, 95), power = 1))
hclplot(diverge_hcl(7, h = c(260, 0), c = 80, l = c(35, 95), power = 1))
```


## Demonstration of statistical graphics:  `demoplot()`

All built-in demos with the same `sequential_hcl(5, "Heat")` palette

```{r echo = FALSE, fig.width = 8.5, fig.height = 8.5, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(3, 3))
cl <- sequential_hcl(5, "Heat")
for (i in c("map", "heatmap", "scatter", "spine", "bar", "pie", "perspective", "mosaic", "lines")) {
  demoplot(cl, type = i)
  mtext(side = 3, text = i, line = 1.5)
}
```

Qualitative palettes: light pastel colors for shading areas (pie)
and darker colorful palettes for points or lines

```{r fig.width = 8.5, fig.height = 3.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
demoplot(qualitative_hcl(4, "Pastel 1"), type = "pie")
demoplot(qualitative_hcl(4, "Set 2"), type = "scatter")
demoplot(qualitative_hcl(4, "Dark 3"), type = "lines")
```

Sequential palettes: display almost continuous gradients with
strong luminance contrasts (heatmap, perspective) and colorful
sequential palette for spine plot with only a few ordered categories

```{r fig.width = 8.5, fig.height = 3.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
demoplot(sequential_hcl(99, "Purple-Blue"), type = "heatmap")
demoplot(sequential_hcl(99, "Reds"), type = "perspective")
demoplot(sequential_hcl(4, "Viridis"), type = "spine")
```

Diverging palettes: display almost continuous gradient with
strong luminance contrast bringing out the extremes (map),
more colorful palette with lower luminance contrasts for displays
with fewer colors (mosaic, bar)

```{r fig.width = 8.5, fig.height = 3.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
demoplot(diverging_hcl(99, "Tropic", power = 2.5), type = "map")
demoplot(diverging_hcl(5, "Green-Orange"), type = "mosaic")
demoplot(diverging_hcl(5, "Blue-Red 2"), type = "bar")
```

Some palettes that work well on black backgrounds

```{r fig.width = 8.5, fig.height = 6.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(2, 3), bg = "black")
demoplot(sequential_hcl(9, "Oslo"), "heatmap")
demoplot(sequential_hcl(9, "Turku"), "heatmap")
demoplot(sequential_hcl(9, "Inferno", rev = TRUE), "heatmap")
demoplot(qualitative_hcl(9, "Set 2"), "lines")
demoplot(diverging_hcl(9, "Berlin"), "scatter")
demoplot(diverging_hcl(9, "Cyan-Magenta", l2 = 20), "lines")
```



# Color vision deficiency emulation

`simulate_cvd()`

* `deutan()`
* `protan()`
* `tritan()`

`deutanomaly_cvd`
`protanomaly_cvd`
`tritanomaly_cvd`


# Color manipulation and utilities

* `desaturate()`
* `darken()`
* `lighten()`
* `max_chroma()`
* `mixcolor()`


# Color apps (shiny and Tcl/Tk)

`choose_palette()`
`hclwizard()`
`hcl_wizard()`

`choose_color()`
`hcl_color_picker()`

`cvd_emulator()`
`interpolate_cvd_transform()`


# HCL-based color scales for ggplot2

All scales are called via the scheme `scale_<aesthetic>_<datatype>_<colorscale>()`,
where `<aesthetic>` is the name of the aesthetic (`fill`, `color`, `colour`),
`<datatype>` is the type of the variable plotted (`discrete` or `continuous`) and
`<colorscale>` sets the type of the color scale used
(`qualitative`, `sequential`, `diverging`, `divergingx`).

Thus, the discrete scales for fill aesthetics are:

* `scale_fill_discrete_qualitative()`
* `scale_fill_discrete_sequential()`
* `scale_fill_discrete_diverging()`
* `scale_fill_discrete_divergingx()`

And the corresponding continuous scales are:

* `scale_fill_continuous_qualitative()`
* `scale_fill_continuous_sequential()`
* `scale_fill_continuous_diverging()`
* `scale_fill_continuous_divergingx()`

Analogous functions for color aesthetics just replace `fill` with `color` (or equivalently `colour`).

**More details:** [HCL-based color scales for _ggplot2_](ggplot2_color_scales.html).
