---
title: "Color Manipulation and Utilities"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: color.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Color Manipulation and Utilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{colorspace}
  %\VignetteKeywords{desaturation, lighten, darken, color mixing}
  %\VignettePackage{colorspace}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("colorspace")
```

## Overview

The _colorspace_ package provides several color manipulation utilities that are
useful for creating, assessing, or transforming color palettes, namely:

* `desaturate()`: Desaturate colors by chroma removal in HCL space.
* `darken()` and `lighten()`: Algorithmically lighten or darken colors in HCL and/or HLS space.
* `max_chroma()`: Compute maximum chroma for given hue and luminance in HCL space.
* `mixcolor()`: Additively mix two colors by computing their convex combination.


## Desaturation in HCL space

Desaturation should map a given color to the gray with the same "brightness".
In principle, any perceptually-based color model (HCL, HLS, HSV, ...) could be
employed for this but HCL works particularly well because its coordinates
capture the perceptual properties better than most other color models.

The `desaturate()` function converts any given hex color code or named
R color to the corresponding HCL coordinates and sets the chroma to zero.
Thus, only the luminance matters which captures the "brightness" mentioned above.
Finally, the resulting HCL coordinates are transformed back to hex color codes
for use in R.

For illustration, a few simple examples are presented below. More examples in the
context of palettes for statistical graphics are discussed along with the
[color vision deficiency](color_vision_deficiency.html) article.

First, `desaturate()` is used to desaturate a vector of R color names:

```{r desaturate-named}
desaturate(c("white", "orange", "blue", "black"))
```

Notice that the hex codes corresponding to three coordinates in sRGB space
are always the same, indicating gray colors.

Analogously, hex color codes can also be transformed - in this case RGB rainbow
colors from the base R function `rainbow()`:

```{r desaturate-rainbow}
rainbow(3)
desaturate(rainbow(3))
```

Even this simple example suffices to show that the three RGB rainbow colors have very
different grayscale levels. This deficiency is even clearer when using 
a full color wheel (of colors with hues in [0, 360] degrees). While the RGB
`rainbow()` is very unbalanced the HCL `rainbow_hcl()` (or also `qualitative_hcl()`)
is (by design) balanced with respect to luminance.

```{r desaturate-wheel, fig.height = 6, fig.width = 6, dev = "png"}
wheel <- function(col, radius = 1, ...)
  pie(rep(1, length(col)), col = col, radius = radius, ...) 
par(mar = rep(0.5, 4), mfrow = c(2, 2))
wheel(rainbow(8))
wheel(desaturate(rainbow(8)))
wheel(rainbow_hcl(8))
wheel(desaturate(rainbow_hcl(8)))
```

## Lighten or darken colors

In principle, a similar approach for lightening and darkening colors can be employed as for desaturation
above. The colors can simply be transformed to HCL space and then the luminance can either be decreased
(turning the color darker) or increased (turning it lighter) while preserving the hue and chroma coordinates.

This strategy typically works well for lightening colors, although in some situations the result can be
rather colorful. Conversely, when darkening rather light colors with little chroma, this can result
in rather gray colors.

In these situations, an alternative might be to apply the analogous strategy in HLS space which is frequently
used in HTML style sheets. However, this strategy may also yield colors that are either too gray or too
colorful. A compromise that sometimes works well is to adjust the luminance coordinate in HCL space but
to take the chroma coordinate corresponding to the HLS transformation.

We have found that typically the HCL-based transformation performs best for lightening colors and this is
hence the default in `lighten()`. For darkening colors, the combined strategy often works best and is hence
the default in `darken()`. In either case it is recommended to try the other available strategies in case
the default yields unexpected results.

Regardless of the chosen color space, the adjustment of the `L` component can occur by two methods, relative
(the default) and absolute. For example `L - 100 * amount` is used for absolute darkening, or
`L * (1 - amount)` for relative darkening. See `lighten()` and `darken()` for more details.

For illustration a qualitative palette (Okabe-Ito) is transformed by two levels of both lightening and
darkening, respectively.

```{r lighten-darken, fig.height = 3.5, fig.width = 4.5, dev = "png"}
oi <- c("#61A9D9", "#ADD668", "#E6D152", "#CE6BAF", "#797CBA")
swatchplot(
  "-40%" = lighten(oi, 0.4),
  "-20%" = lighten(oi, 0.2),
  "  0%" = oi,
  " 20%" =  darken(oi, 0.2),
  " 40%" =  darken(oi, 0.4),
  off = c(0, 0)
)
```


## Adjust transparency of colors

Alpha transparency is useful for making colors semi-transparent, e.g., for
overlaying different elements in graphics. An alpha value
(or alpha channel) of 0 (or 00 in hex strings) corresponds to fully transparent and
an alpha value of 1 (or FF in hex strings) corresponds to fully opaque. If a color
hex string in R does not provide an explicit alpha transparency, the
color is assumed to be fully opaque.

The `adjust_transparency()` function can be used to adjust the alpha transparency
of a set of colors. It always returns a hex color specification. This hex color
can have the alpha transparency added/removed/modified depending on the
specification of the argument `alpha`:

* `alpha = NULL`: Returns a hex vector with alpha transparency only if needed.
  Thus, it keeps the alpha transparency for the colors (if any) but only if
  different from opaque.
* `alpha = TRUE`: Returns a hex vector with alpha transparency
  for all colors, using opaque (FF) as the default if missing.
* `alpha = FALSE`: Returns a hex vector without alpha transparency for
  all colors (even if the original colors had non-opaque alpha).
* `alpha` numeric: Returns a hex vector with alpha transparency for
  all colors set to the `alpha` argument (recycled if necessary).

For illustration, the transparency of a single black color is modified to three
alpha levels: fully transparent, semi-transparent, and fully opaque, respectively.
Black can be equivalently specified by name (`"black"`), hex string (`"#000000"`),
or integer position in the palette (`1`).

```{r transparency-black}
adjust_transparency("black",   alpha = c(0, 0.5, 1))
adjust_transparency("#000000", alpha = c(0, 0.5, 1))
adjust_transparency(1,         alpha = c(0, 0.5, 1))
```

Subsequently, different settings of `alpha` are illustrated for adjusting
a vector with three shades of gray, specified by name (`gray`, opaque),
opaque hex string (`"#BEBEBE"`), and semi-transparent hex string (`"#BEBEBE80"`).
Four types of adjustment are shown: only if necessary (`alpha = NULL`),
add (`alpha = TRUE`), remove (`alpha = FALSE`), or modify (`alpha = 0.8`).

```{r transparency-gray}
x <- c("gray", "#BEBEBE", "#BEBEBE80")
adjust_transparency(x, alpha = NULL)
adjust_transparency(x, alpha = TRUE)
adjust_transparency(x, alpha = FALSE)
adjust_transparency(x, alpha = 0.8)
```


## Maximum chroma for given hue and luminance

As the possible combinations of chroma and luminance in HCL space depend on hue,
it is not obvious which trajectories through HCL space are possible prior to
trying a specific HCL coordinate by calling `polarLUV()`. To avoid having to
fix up the color upon conversion to RGB `hex()` color codes, the
`max_chroma()` function computes (approximately) the maximum chroma
possible. 

For illustration we show that for given luminance (here: L = 50) the maximum chroma
varies substantially with hue:

```{r max-chroma-L50}
max_chroma(h = seq(0, 360, by = 60), l = 50)
```

Similarly, maximum chroma also varies substantially across luminance values for
a given hue (here: H = 120, green):

```{r max-chroma-H120}
max_chroma(h = 120, l = seq(0, 100, by = 20))
```

In the plots below more combinations are visualized: In the left panel for maximum
chroma across hues given luminance and in the right panel with increasing luminance
given hue.

```{r max-chroma-plot, echo = FALSE, fig.height = 4.5, fig.width = 10, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
H <- seq(0, 360, by = 10)
L <- seq(30, 90, by = 20)
C <- outer(H, L, max_chroma)

par(mfrow = c(1, 2), xaxs = "i", yaxs = "i", mar = c(5, 4, 1, 5), las = 1)
plot(0, 0, type = "n", xlim = c(0, 360), ylim = c(0, 165),
  xlab = "Hue (H)", ylab = "Maximum chroma (C)")
axis(4, at = C[37, ], labels = paste("L =", L))
for(i in seq_along(L)) {
  lines(H, C[, i])
  points(H, C[, i], col = hex(polarLUV(L[i], C[, i], H)), pch = 19, cex = 1.5, xpd = TRUE)
}

L <- seq(0, 100, by = 5)
H <- c(0, 60, 120, 250, 330)
C <- outer(H, L, max_chroma)

par(xaxs = "i", yaxs = "i", mar = c(5, 4, 1, 1), las = 1)
plot(0, 0, type = "n", xlim = c(0, 175), ylim = c(0, 100),
  xlab = "Maximum chroma (C)", ylab = "Luminance (L)")
for(i in rev(seq_along(H))) {
  lines(C[i, ], L)
  points(C[i, ], L, col = hex(polarLUV(L, C[i, ], H[i])), pch = 19, cex = 1.5, xpd = TRUE)
  j <- which.max(C[i,])
  text(C[i, j], L[j], paste("H =", H[i]), pos = 4, col = hex(polarLUV(L[j], C[i, j], H[i])))
}
points(0, 100, cex = 1.6, xpd = TRUE)
```


## Additive mixing of two colors

In additive color models like `RGB()` or `XYZ()` it can be useful to combine
colors by additive mixing. Below a fully saturated red and green are mixed,
yielding a medium brownish yellow.

```{r mixcolor, fig.height = 1.5, fig.width = 5, dev = "png", dpi = 48}
R <- RGB(1, 0, 0)
G <- RGB(0, 1, 0)
Y <- mixcolor(0.5, R, G)
Y
par(mar = rep(0, 4))
swatchplot(c(hex(R), hex(Y), hex(G)))
```

