---
title: "Palette Visualization and Assessment"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: color.bib
vignette: >
  %\VignetteIndexEntry{Palette Visualization and Assessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{colorspace}
  %\VignetteKeywords{HCL, palettes, statistical graphics, spectrum}
  %\VignettePackage{colorspace}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("colorspace")
```

## Color swatches: `swatchplot()`

swatches of several palette vectors

```{r swatch-hcl, fig.width = 4, fig.height = 2.2, fig.align = "center", dev = "png"}
swatchplot(
  "Hue"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
  "Chroma"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
  "Luminance" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
  off = 0
)
```

swatches of named palette matrices

```{r swatch-brpg, fig.width = 11.5, fig.height = 3.5, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
bprg <- c("Blues", "Purples", "Reds", "Greens")
swatchplot(
  "Single-hue"            = t(sapply(paste(bprg, 2), sequential_hcl, n = 7)),
  "Single-hue (advanced)" = t(sapply(paste(bprg, 3), sequential_hcl, n = 7)),
  "Multi-hue (advanced)"  = t(sapply(bprg,           sequential_hcl, n = 7)),
  nrow = 5
)
```

## HCL (and RGB) spectrum: `specplot()`

spectrum of HCL-based palettes: qualitative/sequential/diverging

```{r spec-qualitative, fig.width = 5, fig.height = 5, fig.align = "center", dev = "png"}
specplot(qualitative_hcl(100, "Set 2"))
```

```{r spec-sequential, fig.width = 5, fig.height = 5, fig.align = "center", dev = "png"}
specplot(sequential_hcl(100, "Blues 2"))
```

```{r spec-diverging, fig.width = 5, fig.height = 5, fig.align = "center", dev = "png"}
specplot(diverging_hcl(100, "Blue-Red"))
```

spectrum of the (in)famous RGB rainbow palette (in both RGB and HCL)

```{r spec-rainbow, fig.width = 5, fig.height = 7.5, fig.align = "center", dev = "png"}
specplot(rainbow(100), rgb = TRUE)
```


## Trajectories in HCL space: `hclplot()`

For qualitative palettes luminance and chroma are fixed, varying only hue

```{r hcl-qualitative, fig.width = 4, fig.height = 4, fig.align = "center", dev = "png"}
hclplot(qualitative_hcl(9, c = 50, l = 70))
```

Single-hue sequential palette (h = 260) with linear trajectory and
advanced single-hue sequential palette with triangular chroma trajectory
(piecewise linear vs. power-transformed).

```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = 260, c = 80, l = c(35, 95), power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95), power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95), power = c(0.8, 1.4)))
```

Multi-hue sequential palette with small hue range and triangular chroma vs.
large hue range and linear chroma trajectory.

Balanced diverging palette constructed from two simple single-hue sequential
palettes (for hues 260/blue and 0/red)

```{r hcl-multi, fig.width = 11.5, fig.height = 4.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = c(260, 220), c = c(50, 75, 0), l = c(30, 95), power = 1))
hclplot(sequential_hcl(7, h = c(260, 60), c = 60, l = c(40, 95), power = 1))
hclplot(diverge_hcl(7, h = c(260, 0), c = 80, l = c(35, 95), power = 1))
```


## Demonstration of statistical graphics:  `demoplot()`

All built-in demos with the same `sequential_hcl(5, "Heat")` palette

```{r demoplot-all, echo = FALSE, fig.width = 8.5, fig.height = 8.5, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(3, 3))
cl <- sequential_hcl(5, "Heat")
for (i in c("map", "heatmap", "scatter", "spine", "bar", "pie", "perspective", "mosaic", "lines")) {
  demoplot(cl, type = i)
  mtext(side = 3, text = i, line = 1.5)
}
```

Qualitative palettes: light pastel colors for shading areas (pie)
and darker colorful palettes for points or lines

```{r demoplot-qualitative, fig.width = 8.5, fig.height = 3.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
demoplot(qualitative_hcl(4, "Pastel 1"), type = "pie")
demoplot(qualitative_hcl(4, "Set 2"), type = "scatter")
demoplot(qualitative_hcl(4, "Dark 3"), type = "lines")
```

Sequential palettes: display almost continuous gradients with
strong luminance contrasts (heatmap, perspective) and colorful
sequential palette for spine plot with only a few ordered categories

```{r demoplot-sequential, fig.width = 8.5, fig.height = 3.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
demoplot(sequential_hcl(99, "Purple-Blue"), type = "heatmap")
demoplot(sequential_hcl(99, "Reds"), type = "perspective")
demoplot(sequential_hcl(4, "Viridis"), type = "spine")
```

Diverging palettes: display almost continuous gradient with
strong luminance contrast bringing out the extremes (map),
more colorful palette with lower luminance contrasts for displays
with fewer colors (mosaic, bar)

```{r demoplot-diverging, fig.width = 8.5, fig.height = 3.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
demoplot(diverging_hcl(99, "Tropic", power = 2.5), type = "map")
demoplot(diverging_hcl(5, "Green-Orange"), type = "mosaic")
demoplot(diverging_hcl(5, "Blue-Red 2"), type = "bar")
```

Some palettes that work well on black backgrounds

```{r demoplot-dark, fig.width = 8.5, fig.height = 6.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(2, 3), bg = "black")
demoplot(sequential_hcl(9, "Oslo"), "heatmap")
demoplot(sequential_hcl(9, "Turku"), "heatmap")
demoplot(sequential_hcl(9, "Inferno", rev = TRUE), "heatmap")
demoplot(qualitative_hcl(9, "Set 2"), "lines")
demoplot(diverging_hcl(9, "Berlin"), "scatter")
demoplot(diverging_hcl(9, "Cyan-Magenta", l2 = 20), "lines")
```
