---
title: "Palette Visualization and Assessment"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: color.bib
vignette: >
  %\VignetteIndexEntry{Palette Visualization and Assessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{colorspace}
  %\VignetteKeywords{HCL, palettes, statistical graphics, spectrum}
  %\VignettePackage{colorspace}
---

```{r preliminaries, echo=FALSE, message=FALSE}
library("colorspace")
```

## Overview

The _colorspace_ provides several visualization functions for depicting
one or more color palettes and their underlying properties. Color palettes
can be visualized by:

* `swatchplot()`: Color swatches.
* `specplot()`: Spectrum of HCL and/or RGB coordinates.
* `hclplot()`: Trajectories in HCL space projections.
* `demoplot()`: Illustrations of typical (and simplified) statistical graphics.


## Color swatches

The function `swatchplot()` is a convenience function for displaying
collections of palettes that can be specified as lists or matrices of
hex color codes. Essentially, the is just a call to the base graphics
`rect()` function but with heuristics for choosing default labels, margins,
spacings, borders, etc. These are selected to work well for `hcl_palettes()`
and might need further tweaking in future versions.

As a first simple example, the properties of the three HCL axes are brought out:

- Hue: Only changing the type of color with hues H = 0, 60, ..., 300 but
  keeping chroma C = 60 and luminance L = 65 fixed.
- Chroma: Only changing chroma from C = 0 (gray) to 100 (colorful) for fixed hue
  H = 0 (red) and luminance L = 65.
- Luminance: Only changing the luminance from L = 25 (dark) to 90 (light) for
  fixed hue H = 260 and chroma C = 25.

```{r swatch-hcl, fig.width = 4, fig.height = 2.2, fig.align = "center", dev = "png"}
swatchplot(
  "Hue"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
  "Chroma"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
  "Luminance" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
  off = 0
)
```

As a somewhat more complex example a `swatchplot()` is created with three matrices
of sequential color palettes of blues, purples, reds, and greens. For all palettes,
luminance increases monotonically to yield a proper sequential palette. However,
the hue and chroma handling is somewhat different to emphasize different parts of the
palette.

- Single-hue: In each palette the hue is fixed and chroma decreases monotonically
  (along with increasing luminance). This is typically sufficient to clearly bring
  out the extreme colors (dark/colorful vs. light gray).
- Single-hue (advanced): The hue is fixed (as above) but the chroma trajectory is
  triangular. Compared to the basic single-hue palette above this allows to better
  distinguish the colors in the middle and not only the extremes.
- Multi-hue (advanced): As in the advanced single-hue palette the chroma trajectory
  is triangular but additionally the hue varies slightly. This can further enhance
  the distinction of colors in the middle of the palette.

```{r swatch-brpg, fig.width = 11.5, fig.height = 3.5, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
bprg <- c("Blues", "Purples", "Reds", "Greens")
swatchplot(
  "Single-hue"            = t(sapply(paste(bprg, 2), sequential_hcl, n = 7)),
  "Single-hue (advanced)" = t(sapply(paste(bprg, 3), sequential_hcl, n = 7)),
  "Multi-hue (advanced)"  = t(sapply(bprg,           sequential_hcl, n = 7)),
  nrow = 5
)
```

## HCL (and RGB) spectrum

As the properties of a palette in terms of the perceptual dimensions _hue_,
_chroma_, and _luminance_ are not always clear from looking just at color
swatches or (statistical) graphics based on these palettes, the `specplot()`
function provides an explicit display for the HCL coordinates associated
with a palette. This can bring out clearly various aspects, e.g.,
whether hue is constant, chroma is monotonic or triangular, and whether
luminance is approximately constant (as in many qualitative palettes),
monotonic (as in sequential palettes), or diverging.

The function first transforms a given color palette to their HCL (`polarLUV()`)
coordinates. As the hues for low-chroma colors are not (or only poorly) identified,
by default a smoothing is applied to the hues. Also, to avoid jumps from 0 to 360 or vice
versa, the hue coordinates are shifted suitably.


By default, the resulting HCL spectrum is visualized by a simple line plot:

- Hue is drawn in red and coordinates are indicated on the axis on the right
  with range [-360, 360].
- Chroma is drawn in green with coordinates on the left axis. The range [0, 100]
  is used unless the palette necessitates higher chroma values.
- Luminance is drawn in blue with coordinates on the left axis in the range [0, 100].

Additionally, a color swatch for the palette is included. Optionally, a
second spectrum for the corresponding RGB palettes can be included. However,
this is usually just of interest for palettes created in RGB space (or simple
transformations thereof).

The illustrations below show how basic qualitative, sequential, and diverging
palettes are constructed in HCL space. In the qualitative "Set 2" palette below, the
hue traverses the entire color "wheel" (from 0 to 360 degrees) while keeping
chroma and luminance (almost) constant (C = 60 and L = 70).

```{r spec-qualitative, fig.width = 5, fig.height = 5, fig.align = "center", dev = "png"}
specplot(qualitative_hcl(100, "Set 2"))
```

Note that due to the restrictions of the HCL color space, some of the green/blue
colors have a slightly smaller maximum chroma resulting in a small dip in the chroma
curve. This is fixed automatically (by default) and is hardly noticable in visualization,
though.

The sequential "Blues 2" palette below employs a single hue (H = 260) and a monotonically
increasing luminance trajectory (from dark to light). Chroma simply decreases monotonically
along with increasing luminance.

```{r spec-sequential, fig.width = 5, fig.height = 5, fig.align = "center", dev = "png"}
specplot(sequential_hcl(100, "Blues 2"))
```

Note that there is a little kink in the hue curve on the right. However, this is a mere
artifact from the hue being poorly identified at very low chroma values.

Finally, the diverging "Blue-Red" palette is depicted below. It simply combines a blue
and a red sequential single-hue palette (as "Blues 2" above). Hue is constant in each "arm"
of the palette and the chroma/luminance trajectories are balanced between both arms.
In the center the palette passes through a light gray (with zero chroma) as the neutral
value.

```{r spec-diverging, fig.width = 5, fig.height = 5, fig.align = "center", dev = "png"}
specplot(diverging_hcl(100, "Blue-Red"))
```

To contrast these well-balanced HCL-based palettes with a poorly-balanced palette, the
spectrum of the (in)famous RGB rainbow palette is depicted in both RGB and HCL space.

```{r spec-rainbow, fig.width = 5, fig.height = 7.5, fig.align = "center", dev = "png"}
specplot(rainbow(100), rgb = TRUE)
```

The RGB spectrum shows that the trajectories are quite simple in RGB space but the
HCL spectrum shows that the result is very unbalanced with shifts and kinks in both
chroma and (more importantly) luminance. This is why this palette is not suitable
for encoding underlying data in statistical graphics. See also the related discussion
of [color vision deficiency](color_vision_deficiency.html).


## Trajectories in HCL space

While the `specplot()` function above works well for bringing out the HCL coordinates
associated with a given palette, it does not bring out how the palette fits into
the HCL space. For example, it is not so clear whether high chroma values are
close to the maximum possible for a given hue. Thus, it cannot be judged so easily
how the parameters of the hue, chroma, and luminance trajectories can be
modified to obtain another palette.

Therefore, the `hclplot()` is another visualization of the HCL coordinates associated
with a palette. It does so by collapsing over one of the coordinates (either the hue H
or the luminance L) and displays a heatmap of colors combining the remaining
two dimensions. The coordinates for the given color palette are highlighted
to bring out its trajectory. In case the hue is really fixed (as in single-hue
sequential palettes) or the luminance is really fixed (as in the qualitative
palettes), collapsing is straightforward. However, when the coordinate that is
collapsed over is actually not constant in the palette, a simple bivariate linear
model is used to capture how the collapsed coordinate varies along with the
two displayed coordinates.

The function `hclplot()` has been designed to work well with the
`hcl_palettes()` in this package. While it is possible to apply
it to other color palettes as well, the results might look weird or confusing
if these palettes are constructed very differently (e.g., as in the highly
saturated base R palettes). To infer what kind of projection it should use
`hclplot()` checks whether luminance is approximately constant (qualitative),
monotonic (sequential) or triangular (diverging).

Thus, for qualitative palettes - where luminance and chroma are fixed - the varying
hue is displayed in a projection onto the hue-chroma plane at a given fixed luminance:

```{r hcl-qualitative, fig.width = 4, fig.height = 4, fig.align = "center", dev = "png"}
hclplot(qualitative_hcl(9, c = 50, l = 70))
```

The display below compares three single-hue sequential palettes by projection
to the luminance-chroma plane for the given fixed hue. In the left panel
the hue 260 is used with a simple linear chroma trajectory. The other two
panels employ a triangular chroma trajectory for hue 245, either with
a piecewise-linear (middle) or power-transformed (right) trajectory.

```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = 260, c = 80, l = c(35, 95), power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95), power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95), power = c(0.8, 1.4)))
```
Note that for H = 260 it is possible to go to dark colors (= low luminance) with
high chroma while this is not possible to the same extent for H = 245. Hence,
chroma has to be decreased when proceeding to the dark low-luminance colors.

Finally, the display below compares two multi-hue sequential palettes along
with a diverging palette.

```{r hcl-multi, fig.width = 11.5, fig.height = 4.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = c(260, 220), c = c(50, 75, 0), l = c(30, 95), power = 1))
hclplot(sequential_hcl(7, h = c(260, 60), c = 60, l = c(40, 95), power = 1))
hclplot(diverging_hcl(7, h = c(260, 0), c = 80, l = c(35, 95), power = 1))
```

The multi-hue palette on the left employs a small
hue range, resulting in a palette of "blues" just with slightly more distinction
of the middle colors in the palette. In contrast, the multi-hue "blue-yellow"
palette in the middle panel uses a large hue range, resulting in more color
contrasts throughout the palette. Finally, the balanced diverging palette
in the right panel is constructed from two simple single-hue sequential
palettes (for hues 260/blue and 0/red) that are completely balanced between
the two "arms" of the palette.

## Demonstration of statistical graphics


To demonstrate how different kinds of color palettes work in different
kinds of statistical displays, `demoplot()` provides a simple convenience
interface to some base graphics with (mostly artificial) data sets.
All types of demos can deal with arbitrarily many colors. However, some
displays are much more suitable for a low number of colors (e.g., the pie
chart) while others work better with more colors (e.g., the heatmap).

First, all built-in demos are displayed with the same heat colors palette:
`sequential_hcl(5, "Heat")`.

```{r demoplot-all, echo = FALSE, fig.width = 8.5, fig.height = 8.5, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(3, 3))
cl <- sequential_hcl(5, "Heat")
for (i in c("map", "heatmap", "scatter", "spine", "bar", "pie", "perspective", "mosaic", "lines")) {
  demoplot(cl, type = i)
  mtext(side = 3, text = i, line = 1.5)
}
```

_Qualitative palettes:_ Light pastel colors typically work better for
shading areas (pie, left) while darker and more colorful palettes are usually
preferred for points (center) or lines (right).

```{r demoplot-qualitative, fig.width = 8.5, fig.height = 3.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
demoplot(qualitative_hcl(4, "Pastel 1"), type = "pie")
demoplot(qualitative_hcl(4, "Set 2"),    type = "scatter")
demoplot(qualitative_hcl(4, "Dark 3"),   type = "lines")
```

_Sequential palettes:_ Heatmaps (left) or perspective plots (right)
often employ almost continuous gradients with strong luminance contrasts.
In contrast, when only a few ordered categories are to be displayed
(e.g., in a spine plot, right) more colorful sequential palettes 
like the viridis palette can be useful.

```{r demoplot-sequential, fig.width = 8.5, fig.height = 3.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
demoplot(sequential_hcl(99, "Purple-Blue"), type = "heatmap")
demoplot(sequential_hcl(99, "Reds"),        type = "perspective")
demoplot(sequential_hcl( 4, "Viridis"),      type = "spine")
```

_Diverging palettes:_ In some displays (such as the map, left),
it is useful to employ an almost continuous gradient with
strong luminance contrast to bring out the extremes. Here, this
contrast is amplified by a larger power transformation emphasizing
the extremes even further. In contrast, when fewer colors are needed
more colorful palettes with lower luminance contrasts can be desired.
This is exemplified by a mosaic (middle) and bar plot (right).

```{r demoplot-diverging, fig.width = 8.5, fig.height = 3.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(1, 3))
demoplot(diverging_hcl(99, "Tropic", power = 2.5), type = "map")
demoplot(diverging_hcl( 5, "Green-Orange"),        type = "mosaic")
demoplot(diverging_hcl( 5, "Blue-Red 2"),          type = "bar")
```

All displays above focused on palettes designed for light/white backgrounds.
Therefore, to conclude, some palettes are highlighted that work well on
dark/black backgrounds.

```{r demoplot-dark, fig.width = 8.5, fig.height = 6.0, fig.align = "center", dev = "png", dpi = 48, out.width = "100%"}
par(mfrow = c(2, 3), bg = "black")
demoplot(sequential_hcl(9, "Oslo"), "heatmap")
demoplot(sequential_hcl(9, "Turku"), "heatmap")
demoplot(sequential_hcl(9, "Inferno", rev = TRUE), "heatmap")
demoplot(qualitative_hcl(9, "Set 2"), "lines")
demoplot(diverging_hcl(9, "Berlin"), "scatter")
demoplot(diverging_hcl(9, "Cyan-Magenta", l2 = 20), "lines")
```
